Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2017-12-09 paper.tex
final check

In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows biochemists to analyze selected tunnels.
%This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
By selecting a tunnel to be analyzed, the users help the planner to focus only on a subset of the whole configuration
space, similarly as in~\cite{denny2018general}.
Consequently, the volume of the configuration space to be searched is decreased, while the relative volume of the narrow passages is increased.

Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
It can be expected that the tunnels in MD simulations without ligands are rather narrow.
However, the tunnels adapt their shape to the passing ligands and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates shrinking of the ligand size.
In our solution we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.

Molecular dynamics simulations of ligands passing through proteins show that the ligands can be highly flexible.
However, ligands cannot form arbitrary conformations. 
Ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
the dimension of the conifguration space is not increased as in~\cite{cortes2010simulating}.
%So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.

@@

Each conformation from $\L$ is tested during the tree expansion. %in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
Using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in a narrow passage, while the packed conformations can easily rotate on the spot.
%In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
%packed conformations can be used.
%\vspace{-3mm}

@@

We presented a novel modification of the RRT planner to compute trajectories of ligands inside selected protein tunnels.
In our solution, we take into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
The planner generates the random samples only along a tunnel computed by Voronoi diagram, and it attempts to expand the configuration
tree using all ligand conformations. 
The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.

>>>>

In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows the biochemists to analyze selected tunnels.
%This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
By selecting a tunnel to be analyzed, the users help the planner to focus only on a subset of the whole configuration
space, similarly as in~\cite{denny2018general}.
Consequently, the volume of the configuration space to be searched is decreased, while the relative volume of the narrow passages is increased.

Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
It can be expected that the tunnels in MD simulations without ligands are rather narrow.
However, the tunnels adapt their shape to the passing ligands and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
In our solution we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.

Molecular dynamics simulations of ligands passing through proteins show that the ligands can be highly flexible.
However, ligands cannot form arbitrary conformations. 
The ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
the dimension of the configuration space is not increased as in~\cite{cortes2010simulating}.
%So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.

@@

Each conformation from $\L$ is tested during the tree expansion. %in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
Using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed, i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in a narrow passage, while the packed conformations can easily rotate on the spot.
%In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
%packed conformations can be used.
%\vspace{-3mm}

@@

We presented a novel modification of the RRT planner to compute trajectories of ligands inside selected protein tunnels.
In our solution, we take into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
The planner generates the random samples only along a tunnel computed by the Voronoi diagram, and it attempts to expand the configuration
tree using all ligand conformations. 
The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
